What Makes Deep Network Representations Good for Visual Recognition?

### 2015-04-07 paper_for_review.tex
v1.3

% Include other packages here, before hyperref.
\usepackage[labelfont=bf, skip=3pt]{caption}
\captionsetup{labelsep=period}
\usepackage{footmisc}

@@

Representations of artificial neurons in deep neural networks have been demonstrated to be highly effective and robust for various visual recognition tasks~\cite{krizhevsky2012imagenet, sermanet2013overfeat, donahue2014decaf}.
Nevertheless, our understanding of them arguably remains limited.
In this paper, we address two important and practical questions: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others. Using methods which support characterizing both first-order and second-order properties of scalar and vector representations, we considered the case study of face pair matching and conducted a large-scale experiment. This experiment examined 200 randomly generated shallow and deep networks, with the results analyzed through a series of measures designed to quantify the differences in representations. 
We conclude through this principled analysis that deep representations are quantitatively better than shallow representations, and overall 71\% of the variance in the performance of deep networks can be explained with our measures, both of which are confirmed using statistical tests for the first time.

%(\ie optimal stimulus) (\ie invariance and selectivity paths)
%use X as example of deep networks
%concern about deep networks; how alleviated via this work

\end{abstract}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90\columnwidth]{Figs/e_fig7s_compact-crop.pdf} 
\end{center}
\caption{{\bf What makes a deep network good?}
In existing work, researchers have mostly tried to produce one ``best'' network architecture (\ie a hyperparameter set, either through manual tuning or automated hyperparameter search), which is subsequently evaluated on a particular benchmark. However, as exemplified in this figure, when considering a larger population of networks \cite{cox2011beyond}, it is not fully clear to us why in practice deep networks (\eg two or more levels) outperform shallow networks (\eg one-level), and why certain networks are better than the others, given their broad distribution of performance on a task like face pair matching for the LFW-a data set (shown above)~\cite{LFWTech, wolf2011effective}.
%However, it is not fully clear that in practice (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
%As exemplified in the figure, deep (\ie two-level) networks not only on average outperform shallow (\ie one-level) networks \cite{cox2011beyond}, but also exhibit a wide range of performance differences on LFW-a face recognition \cite{LFWTech, wolf2011effective}.
} % (\eg optimal stimulus). %and (sometimes) characterized its properties
\label{fig:perf}
\end{figure}

Although deep neural networks have been successfully pushing the boundaries of computer vision by breaking the records of ImageNet Challenges (ILSVRC) year after year~\cite{krizhevsky2012imagenet, sermanet2013overfeat, szegedy2014going}, our understanding of the properties of these networks arguably remains limited, since we cannot yet fully answer the question ``what makes a deep network good?'' In this paper, we investigate two important facets of this question: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than others (Fig.~\ref{fig:perf}). Specifically, from the point of view that successful networks are forming ``better'' representations of visual inputs that are more separable to the final-stage linear classifiers, we want to identify the crucial properties that are required for forming (or learning) better representations.

@@

%first order, second order, flexibility
{\bf Existing work.} 
Given an artificial neural network $f$, the study of how $f$ represents a high dimensional stimulus\footnote{\expstimdim} ${\bf{x}} \in \mathbb{R}^N$ as a representation 
${\bf{r}} = f\left( {\bf{x}} \right)$ in existing work mainly focused on the first-order (\ie linear) characterization of $f$---the analysis of optimal stimuli, stimuli which lead to the strongest responses of given neurons.
For single-level convolutional networks using squared pooling, \ie $f({\bf{x}}) = \left\| {\bf{K}} \otimes {\bf{x}} \right\|_{F}$, the optimal stimulus was shown to be analytically approximable \cite{saxe2011random} using a Gabor-like filter, whether the convolution kernel $\bf{K}$ is structural or random.
For multi-level convolutional networks, the optimal stimuli of neurons from various depths were numerically approximated \cite{ngiam2010tiled, le2012building, zeiler2014visualizing, simonyan2013deep} to qualitatively show neurons in deeper layers are tuned to more complex visual patterns.
The generalized concept of optimal stimulus in considering groups of neurons can also be adopted to study the representations by inverting them \cite{mahendran2014understanding}.

In addition to the first-order characterization, the analysis of invariance and selectivity properties of neurons (\eg changes in stimuli which given neurons are least and most sensitive to, how small or large the changes are, \etc) was adopted in a few works. This can be viewed as the second-order characterization of $f$, and parametric deformation in stimuli was the most commonly used setup.
For example, translation, rotation, scaling, \etc of the stimulus were used in~\cite{goodfellow2009measuring, zeiler2014visualizing} to test the invariance and selectivity of neurons.
Similar deformations were also adopted in \cite{lenc2014understanding}, but extended to analyze more advanced properties (\ie equivariance and equivalence) of groups of neurons.
More generalized setups that, \eg, utilize local quadratic information (\ie the Hessian) or ``global'' numerical searches in the stimulus space, even though they do exist, are less explored in the context of artificial neural network studies.
For quadratic networks, \ie $f\left({\bf{x}}\right) = \frac{1}{2}{\bf{x}}^{T}{\bf{Qx}}+{\bf{L}}^{T}{\bf{x}}+c$, through eigendecomposing the quadratic term $\bf{Q}$ (\ie the Hessian), local invariance and selectivity directions can be analytically derived \cite{berkes2006analysis}, which correspond to eigenvectors of the least and most negative eigenvalues.
This method was also generalized onto multi-layer convolutional networks by numerically estimating the Hessian \cite{ngiam2010tiled}.
For multi-layer networks, characterization of the ``global'' invariance of single neurons can also be achieved by numerically searching for non-local (\ie farther to the optimal stimuli) solutions \cite{erhan2010understanding}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=1.0\columnwidth]{Figs/methods.jpg} 
\end{center}
\caption{{\bf Representation characterization methods.}
On the spherical (\ie constant energy) constraint of the $N$ dimensional stimulus space, similar to previous work, the first-order property (\ie optimal stimulus $\hat{\bf{x}}$) of a representation $\bf{r}$ is first iteratively searched (as the evolution of light gray eclipses) and analyzed.
However, unlike most previous work, the second-order properties (\ie invariant path $\lbrace \bf{x}^{+}_{\delta} \rbrace$ and selectivity path $\lbrace \bf{x}^{-}_{\delta} \rbrace$) at distances $0.1\pi$ to $0.5\pi$ from the optimal stimulus are also searched and analyzed.
Additionally, these methods are capable of characterizing both scalar and vector representations, where the ``fitnesses'' to be optimized correspond to Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1} and Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}, respectively.
In this example, the scalar representation (\ie response) of the given neuron is tuned to a $45^{\circ}$ Gabor filter and invariant to phase changes, but selective to orientation changes.
}
\label{fig:methods}
\end{figure}

@@

To study the representations inside an artificial neural network, methods adopted in this work include: (1) first-order (\ie linear) characterization, or search and analysis of optimal stimuli, and (2) second-order (\ie quadratic) characterization, or search and analysis of invariance and selectivity paths (\ie eigenvectors of the largest and smallest eigenvalues), which are generalized and improved from previous work.
$\bf{r}$ is defined as a scalar representation (\ie $\bf{r} \in \mathbb{R}$) when considering single neurons and a vector representation (\ie ${\bf{r}} \in \mathbb{R}^C$) when considering groups of neurons (with group size $C$).
Our methods support characterizing both scalar (using Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1}, though Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2} can be used as well) and vector (using Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}) representations in a unified way.
The search phases defined via these equations will be introduced in the following subsections, and the analyses of search results using representation measures will be covered in Sec.~\ref{sec:exp}.

@@

While no specific knowledge of the network itself is assumed, we do restrict the space of stimuli considered with a constant energy constraint $\left\| \bf{x} \right\| = E$, where $E$ is set to $1$ in all mathematical derivations for simplicity, but set to the average energy of task-related images in our experiments.
Since neurons can often be modulated by stimulus contrasts, which is less interesting when considering pattern selectivities, limiting the stimulus space in this manner effectively avoids degenerate solutions that simply maximize stimulus contrasts, and reduces the range of possible stimuli to consider.

@@

\newcommand{\expdiff}{The way the simple linear constraint is constructed to enforce the exploration of a larger extent of the fitness landscape is also one of the main differences compared to \cite{erhan2010understanding}.}

With respect to the optimal stimulus $\hat{\bf{x}}$, the searches of invariance and selectivity paths, which consist of sets of invariant stimuli $\left\lbrace \bf{x}^{+}_{\delta} \right\rbrace$ and selective stimuli $\left\lbrace \bf{x}^{-}_{\delta} \right\rbrace$ respectively, are formulated as
\begin{align}
\bf{x}^{+}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\max} f\left(\bf{x}_{\delta}\right) \label{eq:I1} \\
&= \underset{\bf{x}_{\delta}}{\arg\max} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:I2} \\
\bf{x}^{-}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\min} f\left(\bf{x}_{\delta}\right) \label{eq:S1} \\
&= \underset{\bf{x}_{\delta}}{\arg\min} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:S2}
\end{align}
in which $0 < \delta \le \frac{\pi}{2}$, subject to $\left\| \bf{x}_{\delta} \right\| = 1$ and $\langle \bf{x}_{\delta} , \hat{\bf{x}} \rangle = \cos\left(\delta\right)$. 
As visualized in Fig.~\ref{fig:methods}, the invariance path characterizes the $N$ dimensional curve that leads toward (or ideally, maintains at) a fitness as high as possible while moving away from the optimal stimulus, and the selectivity path characterizes the curve that leads toward fitness as low as possible while moving away.
The search process is implemented as multiple runs of maximization and minimization on discretized $\delta \in \left\lbrace 0.1\pi, 0.2\pi, 0.3\pi, 0.4\pi, 0.5\pi\right\rbrace$ as the circular distance constraints shown in Fig.~\ref{fig:methods}, where each run is initialized with the result from the previous run (and the $0.1\pi$ run directly with optimal stimulus $\hat{\bf{x}}$) to increase the path continuity and searching speed\footnote{\expdiff}.
In this work, the distance constraint $\delta$ only goes up to $\frac{\pi}{2}$, since on the $N$ dimensional sphere, stimuli that fall in $\frac{\pi}{2} < \delta \le \pi$ are simply ``negatives'' of those in $0 < \delta \le \frac{\pi}{2}$, which are of less uniqueness and interest in general. 
The same process can also be performed with respect to a reference stimulus $\tilde{\bf{x}}$, especially for vector representations, where the invariance and selectivity of the ``imaginary neuron'' are again characterized equivalently.

@@

For our experiments, a simplified variety of deep convolutional neural networks---the Spatio-Temporal Hierarchical Object Representation (STHOR) model \cite{pinto2009high, sthor}---{was} adopted.
Like other deep networks, it consists of the standard cascade of convolution, nonlinear activation, pooling layers, \etc (which together define a single ``level'' in this work). However, its convolution kernel weights are  randomly assigned rather than trained.
Though not matching the performances of deep networks trained on massive quantities of data \cite{krizhevsky2012imagenet}, it can in fact be surprisingly competitive for a variety of tasks, especially in the regime where large quantities of training samples are not readily available \cite{pinto2009high, cox2011beyond, viglarge}.
Most importantly, since it does not require training (except for the linear SVM on top of the representations extracted), a large number of networks with diverse hyperparameters can be rapidly generated and compared, which is particularly useful in this study where relative, instead of absolute task performances, are of primary interest. 
As one might be concerned that comparing randomly weighted networks will not tell much about representations, since they might as well be random, our results will suggest otherwise.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.80\textwidth]{Figs/pic1.jpg} 
\end{center}
\caption{{\bf Visualization of properties of shallow (A--D) and deep scalar representations (E--H).}
Color of the boarder of an image indicates the fitness of the representation (\ie response of the neuron) driven by the image, with color map following the definition in Fig.~\ref{fig:methods}.
A and E show the search trajectories of \emph{optimal stimuli} of all 32 neurons from the best performing shallow and best performing deep networks, respectively, where A.1 and E.1 show the initial $1/f$ random stimuli, A.5 and E.5 the resultant optimal stimuli, and A.2--4 and E.2--4 three intermediate steps corresponding to the three largest curvatures in the fitness histories.
B and F show the \emph{invariance paths} of shallow and deep neurons, starting from the corresponding optimal stimuli show in A.5 and E.5, moving from $0.1\pi$ (B.1 and F.1) to $0.5\pi$ (B.5 and F.5) away accordingly.
C and G show the \emph{selectivity paths} of shallow and deep neurons, with definitions following B and F.
D and H show the \emph{invariance subspaces} sampled $0.1\pi$ away from the optimal stimuli, with 5 out of 20 results randomly selected.
Comparatively, \emph{deep neurons show more complex and diverse representational properties} than shallow neurons do.
} % do visually
\label{fig:allrep}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=0.75\columnwidth]{Figs/fig4.pdf} 
\end{center}
\caption{{\bf Fitness-distance analyses of shallow (A) and deep scalar representations (B).}
Fitnesses of invariance (red dots, \ie $\left\lbrace f\left(\bf{x}^{+}_{\delta}\right) \right\rbrace$) and selectivity paths (blue dots, \ie $\left\lbrace f\left(\bf{x}^{-}_{\delta}\right) \right\rbrace$) of all 3,200 shallow and 3,200 deep neurons are shown for comparison\protect\footnotemark.
Although both types of neurons show invariance, shallow neurons do not posses selectivity, since their selectivity paths overlap with the zero-selectivity baseline.
In other words, \emph{only deep neurons are selective}.
} %, and random walks (green dots, \ie $\left\lbrace f\left({\bf{x}}^{R}_{\delta}\right) \right\rbrace$)
%the dashed line, 
\label{fig:fda}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.80\columnwidth]{Figs/fig5.pdf} 
\end{center}
\caption{{\bf Comparison of shallow and deep scalar representation measures.}
Representation measures from left to right are, optimal stimulus spectral complexity (OSSC), invariance path potential (INPP), selectivity path potential (SLPP), and invariance subspace capacity (INSC).
Sensitivity indexes ($d'$) between distributions are $1.63$, $0.42$, $4.73$, and $3.67$, and differences between means of distributions are all significant (***), which suggest \emph{deep representations are quantitatively and significantly better than shallow representations}.
} %, in which data points from shallow networks (left distributions) and deep networks (right distributions) are shown side by side
\label{fig:pair}
\end{figure}

{\bf Optimal stimulus spectral complexity.} Figure \ref{fig:allrep}A and \ref{fig:allrep}E illustrate the trajectories of optimal stimulus searches of 32 top-layer neurons from the best performing shallow network and 32 from the best performing deep network respectively.
To further increase the searching speed, the initial stimulus {was} selected from 1000 ${1}/{f^{\alpha}}$ random stimuli with the best fitness, where $\alpha \in \left\lbrace -4,-3,-2,-1,0 \right\rbrace$, without overly sacrificing the nature of random initialization.
From the results, first, the response landscapes of deep neurons are noticeably more complex than those of shallow neurons, as a large fraction of searching trajectories for deep neurons actually go down then up before reaching the (local) optima, which also suggests the numerical solver can handle non-convexity reasonably well.
Second, the optimal stimuli of deep neurons are also relatively more complex (\ie consisting of more textural components) than those of shallow neurons (which are more uni-textured or Gabor-like).
To quantify such differences, the spectral complexity measure, which is estimated through the $L^{1}$ norm of the (2 dimensional) Fourier power spectrum of optimal stimulus, \ie $\left\| \mathcal{F}(\hat{\bf{x}}) \right\|_{1}$, is adopted, where a higher value suggests higher non-sparsity (\ie more spectral components required to represent the visual stimuli).

@@

{\bf Invariance and selectivity path potential.} Figure \ref{fig:allrep}B/C and \ref{fig:allrep}F/G illustrate the invariance and selectivity paths on the same sets of shallow and deep neurons as in \ref{fig:allrep}A and \ref{fig:allrep}D.
While invariance paths are mostly phase changes and selectivity paths are leading toward meaningless noises (all at the same falloff rate) for shallow neurons, both types of paths consist of sophisticated shape deformations for deep neurons.
As further demonstrated in the fitness-distance diagrams \cite{jones1995fitness} (Fig.~\ref{fig:fda}), on average, although shallow neurons have good invariance (\ie red curve stays high), they do not have any selectivity (\ie blue curve drops slow), while deep neurons show both good invariance and selectivity (\ie red curve stays high and blue curve drops fast).
For scalar representations, we can define the invariance (and selectivity, similarly) path potential as the area sandwiched between the invariance curve and the cosine baseline curve (\ie the invariance and selectivity curves of a single inner-product neuron\footnote{\defbaseline}, or the ``baseline'' of neural network types) on the fitness-distance diagram, \ie $\int_{0}^{\frac{\pi}{2}}{\left| \cos^{-1}\left(f\left( \bf{x}^{+}_{\delta} \right)\right) - \delta \right|}\mathrm{d}\delta \mathbin{/} {\frac{\pi}{2}}$ as normalized in the $\cos^{-1}$ domain, to measure how invariant (and selective) a neuron is compared to the baseline.

>>>>

% Include other packages here, before hyperref.
%\usepackage{float}
\usepackage[font=small, labelfont=bf, skip=3pt]{caption} %
\captionsetup{labelsep=period}
\usepackage{footmisc}

@@

Representations of artificial neurons in deep neural networks have been demonstrated to be highly effective and robust for various visual recognition tasks \cite{krizhevsky2012imagenet, sermanet2013overfeat, donahue2014decaf}.
Nevertheless, our understanding of them arguably remains limited.
In this paper, we address two important and practical questions: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
Using methods which support characterizing both first- and second-order properties of scalar and vector representations, we considered the case study of face pair matching and conducted a large-scale experiment.
The experiment examined 200 randomly generated shallow and deep networks, with the results analyzed via a series of measures designed to quantify the differences in representations. 
We conclude through this principled analysis that deep representations are quantitatively better than shallow representations, and overall 71\% of the variance in the performance of deep networks can be explained with our measures, both of which are confirmed using statistical tests for the first time.

%(\ie optimal stimulus) (\ie invariance and selectivity paths)
%use X as example of deep networks
%concern about deep networks; how alleviated via this work
\end{abstract}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.85\columnwidth]{Figs/e_fig7s_compact-crop.pdf} 
\end{center}
\caption{{\bf What makes a deep network good?}
In existing work, researchers have mostly tried to produce one ``best'' network architecture (\ie a hyperparameter set, through manual tuning or automated hyperparameter search), which is subsequently evaluated on a particular benchmark. However, when considering a larger population of networks \cite{cox2011beyond}, it is not fully clear why in practice deep networks (\eg two or more levels) outperform shallow networks (\eg one level), and why certain networks are better than the others, given their broad distribution of performance on a task like face pair matching for the LFW-a dataset \cite{LFWTech, wolf2011effective} as shown above.
%However, it is not fully clear that in practice (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
%As exemplified in the figure, deep (\ie two-level) networks not only on average outperform shallow (\ie one-level) networks \cite{cox2011beyond}, but also exhibit a wide range of performance differences on LFW-a face recognition \cite{LFWTech, wolf2011effective}.
%as exemplified in this figure,
} % (\eg optimal stimulus). %and (sometimes) characterized its properties
\label{fig:perf}
\end{figure}

Although deep neural networks have been successfully pushing the boundaries of computer vision by breaking the records of ImageNet Challenges (ILSVRC) \cite{russakovsky} year after year \cite{krizhevsky2012imagenet, sermanet2013overfeat, szegedy2014going}, our understanding of the properties of these networks arguably remains limited, since we cannot yet fully answer the question ``what makes a deep network good?'' In this paper, we investigate two important facets of this question: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than others (Fig.~\ref{fig:perf}). Specifically, from the point of view that successful networks are forming ``better'' representations of visual inputs that are more separable to the final-stage linear classifiers, we want to identify the crucial properties that are required for forming (or learning) better representations.

@@

%first order, second order, flexibility
{\bf Existing work.} 
When given an artificial neural network $f$, the study of how $f$ represents a high dimensional stimulus\footnote{\expstimdim} ${\bf{x}} \in \mathbb{R}^N$ as a representation 
${\bf{r}} = f\left( {\bf{x}} \right)$ in existing work mainly focused on the first-order (\ie linear) characterization of $f$---the analysis of optimal stimuli, stimuli which lead to the strongest responses of given neurons.
For single-level convolutional networks using squared pooling, \ie $f({\bf{x}}) = \left\| {\bf{K}} \otimes {\bf{x}} \right\|_{F}$, the optimal stimulus was shown to be analytically approximable \cite{saxe2011random} using a Gabor-like filter, whether the convolution kernel $\bf{K}$ is structural or random.
For multi-level convolutional networks, the optimal stimuli of neurons from various depths were numerically approximated \cite{ngiam2010tiled, le2012building, zeiler2014visualizing, simonyan2013deep} to qualitatively show neurons in deeper layers are tuned to more complex visual patterns.
The generalized concept of optimal stimulus in considering groups of neurons can also be adopted to study the representations by inverting them \cite{mahendran2014understanding}.

In addition to the first-order characterization, the analysis of invariance and selectivity properties of neurons (\eg changes in stimuli which given neurons are least and most sensitive to, how small or large the changes are, \etc) was adopted in a few works. This can be viewed as the second-order characterization of $f$, and parametric deformation in stimuli was the most commonly used setup.
For example, translation, rotation, scaling, \etc of the stimulus were used in \cite{goodfellow2009measuring, zeiler2014visualizing} to test the invariance and selectivity of neurons.
Similar deformations were also adopted in \cite{lenc2014understanding}, but extended to analyze more advanced properties (\ie equivariance and equivalence) of groups of neurons.
More generalized setups that, \eg, utilize local quadratic information (\ie the Hessian) or ``global'' numerical searches in the stimulus space, even though they do exist, are less explored in the context of artificial neural network studies.
For quadratic networks, \ie $f\left({\bf{x}}\right) = \frac{1}{2}{\bf{x}}^{T}{\bf{Qx}}+{\bf{L}}^{T}{\bf{x}}+c$, through eigendecomposing the quadratic term $\bf{Q}$ (\ie the Hessian), local invariance and selectivity directions can be analytically derived \cite{berkes2006analysis}, which correspond to eigenvectors of the least and most negative eigenvalues.
This method was also generalized onto multi-layer convolutional networks by numerically estimating the Hessian \cite{ngiam2010tiled}.
For multi-layer networks, characterization of the ``global'' invariance of single neurons can also be achieved by numerically searching for non-local (\ie farther to the optimal stimuli) solutions \cite{erhan2010understanding}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90\columnwidth]{Figs/methods.jpg} 
\end{center}
\caption{{\bf Methods of representation characterization.}
On the spherical (\ie constant energy) constraint of the $N$ dimensional stimulus space, similar to previous work, the first-order property (\ie optimal stimulus $\hat{\bf{x}}$) of a representation $\bf{r}$ is first iteratively searched (as the evolution of gray eclipses) and analyzed.
However, unlike most previous work, the second-order properties (\ie invariant path $\lbrace \bf{x}^{+}_{\delta} \rbrace$ and selectivity path $\lbrace \bf{x}^{-}_{\delta} \rbrace$) at distances $0.1\pi$ to $0.5\pi$ from the optimal stimulus are also searched and analyzed.
Additionally, these methods can characterize both scalar and vector representations, in which the ``fitnesses'' to be optimized correspond to Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1} and Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}, respectively.
In this example, the scalar representation (\ie response) of the given neuron is tuned to a $45^{\circ}$ Gabor filter and invariant to phase changes, but selective to orientation changes.
}
\label{fig:methods}
\end{figure}

@@

To study the representations inside an artificial neural network, methods adopted in this work include: (1) first-order (\ie linear) characterization, or search and analysis of optimal stimuli, and (2) second-order (\ie quadratic) characterization, or search and analysis of invariance and selectivity paths (\ie eigenvectors of the largest and smallest eigenvalues), which are generalized and improved from previous work.
In this paper, $\bf{r}$ is defined as a scalar representation (\ie $\bf{r} \in \mathbb{R}$) when considering single neurons and a vector representation (\ie ${\bf{r}} \in \mathbb{R}^C$) when considering groups of neurons (with group size $C$).
Our methods support characterizing both scalar (using Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1}, though Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2} can be used as well) and vector (using Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}) representations in a unified way.
The search phases defined via these equations will be introduced in the following subsections, and the analyses of search results using representation measures will be covered in Sec.~\ref{sec:exp}.

@@

While no specific knowledge of the network itself is assumed, we do restrict the space of stimuli considered with a constant energy constraint $\left\| \bf{x} \right\| = E$, where $E$ in all mathematical derivations is set to $1$ for simplicity, but in our experiments, set to the average energy of task-related images.
Since neurons can often be modulated by stimulus contrasts, which is less interesting when considering pattern selectivities, limiting the stimulus space in this manner effectively avoids degenerate solutions that simply maximize stimulus contrasts, and reduces the range of possible stimuli to consider.

@@

\newcommand{\expdiff}{The way we constructed the simple linear constraints to enforce the exploration of the landscape is also one of the main differences compared to methods in \cite{erhan2010understanding}.}
%to enforce the exploration of a larger extent of the fitness landscape

With respect to the optimal stimulus $\hat{\bf{x}}$, the searches of invariance and selectivity paths, which consist of sets of invariant stimuli $\left\lbrace \bf{x}^{+}_{\delta} \right\rbrace$ and selective stimuli $\left\lbrace \bf{x}^{-}_{\delta} \right\rbrace$ respectively, are formulated as
\begin{align}
\bf{x}^{+}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\max} f\left(\bf{x}_{\delta}\right) \label{eq:I1} \\
&= \underset{\bf{x}_{\delta}}{\arg\max} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:I2} \\
\bf{x}^{-}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\min} f\left(\bf{x}_{\delta}\right) \label{eq:S1} \\
&= \underset{\bf{x}_{\delta}}{\arg\min} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:S2}
\end{align}
in which $0 < \delta \le \frac{\pi}{2}$, subject to $\left\| \bf{x}_{\delta} \right\| = 1$ and $\langle {\bf{x}}_{\delta} , \hat{{\bf{x}}} \rangle = \cos\left(\delta\right)$. 
As visualized in Fig.~\ref{fig:methods}, the invariance path characterizes the $N$ dimensional curve that leads toward (or ideally, maintains at a) fitness as high as possible while moving away from the optimal stimulus, and the selectivity path characterizes the curve that leads toward fitness as low as possible while moving away.
The search process is implemented as multiple runs of maximization and minimization on discretized $\delta \in \left\lbrace 0.1\pi, 0.2\pi, 0.3\pi, 0.4\pi, 0.5\pi\right\rbrace$ as the circular distance constraints shown in Fig.~\ref{fig:methods}, where each run is initialized with the result from the previous run (and the $0.1\pi$ run directly with optimal stimulus $\hat{\bf{x}}$) to increase the path continuity and searching speed\footnote{\expdiff}.
In this work, the distance constraint $\delta$ only goes up to $\frac{\pi}{2}$, since on the $N$ dimensional sphere, stimuli that fall in $\frac{\pi}{2} < \delta \le \pi$ are simply ``negatives'' of those in $0 < \delta \le \frac{\pi}{2}$, which are of less uniqueness and interest in general. 
The same process can also be performed with respect to a reference stimulus $\tilde{\bf{x}}$, especially for vector representations, where the invariance and selectivity of the aforementioned ``imaginary neuron'' are again characterized equivalently.

@@

For our experiments, a simplified variety of deep convolutional neural networks---the Spatio-Temporal Hierarchical Object Representation (STHOR) model \cite{pinto2009high, sthor}---{was} adopted.
Like other deep networks, it consists of the standard cascade of convolution, nonlinear activation, pooling layers, \etc (which together define a single ``level'' in this work). However, its convolution kernel weights are  randomly assigned rather than trained.
Though not matching the performances of deep networks trained on massive quantities of data \cite{krizhevsky2012imagenet}, it can in fact be surprisingly competitive for a variety of tasks, especially in the regime where large quantities of training samples are not readily available \cite{pinto2009high, cox2011beyond, viglarge}.
Most importantly, since it does not require training (except for the linear SVM on top of the representations extracted), a large number of networks with diverse hyperparameters can be rapidly generated and compared, which is particularly useful in this study where relative, instead of absolute task performances, are of primary interest. 
One might be concerned that comparing randomly weighted networks will not tell much about representations; however, our results will suggest otherwise.

%(since they might as well be random)

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.845\textwidth]{Figs/pic1.jpg} 
\end{center}
\caption{{\bf Visualization of properties of shallow (A--D) and deep scalar representations (E--H).}
Color of the boarder of an image indicates the fitness of the representation (\ie response of the neuron) driven by the image, with color map following the definition in Fig.~\ref{fig:methods}.
A and E show the search trajectories of \emph{optimal stimuli} of all 32 neurons from the best performing shallow and best performing deep networks, respectively, where A.1 and E.1 show the initial $1/f$ random stimuli, A.5 and E.5 the resultant optimal stimuli, and A.2--4 and E.2--4 three intermediate steps corresponding to the three largest curvatures in the fitness histories.
B and F show the \emph{invariance paths} of shallow and deep neurons, starting from the corresponding optimal stimuli show in A.5 and E.5, moving from $0.1\pi$ (B.1 and F.1) to $0.5\pi$ (B.5 and F.5) away accordingly.
C and G show the \emph{selectivity paths} of shallow and deep neurons, with definitions following B and F.
D and H show the \emph{invariance subspaces} sampled $0.1\pi$ away from the optimal stimuli, with 5 out of 20 results randomly selected.
Comparatively, \emph{deep neurons show more complex and diverse representational properties} than shallow neurons do.
} % do visually
\label{fig:allrep}
\end{figure*}

{\bf Optimal stimulus spectral complexity.} Figure \ref{fig:allrep}A and \ref{fig:allrep}E illustrate the trajectories of optimal stimulus searches of 32 top-layer neurons from the best performing shallow network and 32 from the best performing deep network respectively.
To further increase the searching speed, the initial stimulus {was} selected from 1000 ${1}/{f^{\alpha}}$ random stimuli with the best fitness, where $\alpha \in \left\lbrace -4,-3,-2,-1,0 \right\rbrace$, without overly sacrificing the nature of random initialization.
From the results, first, the response landscapes of deep neurons are noticeably more complex than those of shallow neurons, as a large fraction of search trajectories for deep neurons actually go down then up before reaching the (local) optima, which also suggests the numerical solver can handle non-convexity reasonably well.
Second, the optimal stimuli of deep neurons are also relatively more complex (\ie consisting of more textural components) than those of shallow neurons (which are more uni-textured or Gabor-like).
To quantify such differences, the spectral complexity measure, which is estimated through the $L^{1}$ norm of the two dimensional Fourier power spectrum of optimal stimulus, \ie $\left\| \mathcal{F}(\hat{\bf{x}}) \right\|_{1}$, is adopted, where a higher value suggests higher non-sparsity (\ie more spectral components required to represent the visual stimuli).

\begin{figure}
\begin{center}
\includegraphics[width=0.75\columnwidth]{Figs/fig4.pdf} 
\end{center}
\caption{{\bf Fitness-distance analyses of shallow (A) and deep scalar representations (B).}
Fitnesses of invariance (red dots, \ie $\left\lbrace f\left(\bf{x}^{+}_{\delta}\right) \right\rbrace$) and selectivity paths (blue dots, \ie $\left\lbrace f\left(\bf{x}^{-}_{\delta}\right) \right\rbrace$) of all 3,200 shallow and 3,200 deep neurons are shown for comparison\protect\footnotemark.
Although both types of neurons show invariance, shallow neurons do not posses selectivity, since their selectivity paths overlap with the zero-selectivity baseline, \ie \emph{only deep neurons are selective}.
} %, and random walks (green dots, \ie $\left\lbrace f\left({\bf{x}}^{R}_{\delta}\right) \right\rbrace$)
%the dashed line, 
\label{fig:fda}
\end{figure}

@@

{\bf Invariance and selectivity path potential.} Figure \ref{fig:allrep}B/C and \ref{fig:allrep}F/G illustrate the invariance and selectivity paths on the same sets of shallow and deep neurons as in \ref{fig:allrep}A and \ref{fig:allrep}D.
While invariance paths are mostly phase changes and selectivity paths are leading toward meaningless noises (all at the same falloff rate) for shallow neurons, both types of paths consist of sophisticated shape deformations for deep neurons.
As further demonstrated in the fitness-distance diagrams \cite{jones1995fitness} (Fig.~\ref{fig:fda}), on average, although shallow neurons have good invariance (\ie red curve stays high), they do not have any selectivity (\ie blue curve drops slow), while deep neurons show both good invariance and selectivity (\ie red curve stays high and blue curve drops fast).
For scalar representations, we can define the invariance (and selectivity, similarly) path potential as the area sandwiched between the invariance curve and the baseline curve (\ie the invariance and selectivity curves of a single inner-product neuron\footnote{\defbaseline}, or the ``baseline'' of all neural network types) on the fitness-distance diagram, \ie $\int_{0}^{\frac{\pi}{2}}{\left| \cos^{-1}\left(f\left( \bf{x}^{+}_{\delta} \right)\right) - \delta \right|}\mathrm{d}\delta \mathbin{/} {\frac{\pi}{2}}$ as normalized in the $\cos^{-1}$ domain, to measure how invariant (and selective) a neuron is compared to the baseline.
